Exact Phylodynamics via Structured Markov Genealogy Processes

### 2024-01-17 ms.Rnw
update
rewrite

\documentclass[11pt,reqno,final]{amsart}
\usepackage[round,elide]{natbib}
\input{header}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,decorations,calc,math}

@@

<<prefix,include=FALSE,cache=FALSE,purl=FALSE>>=
prefix <- "smgp"
source("setup.R")
@
<<packages,include=FALSE,cache=FALSE>>=
library(tidyverse)
library(ggtree)
library(pomp)
library(cowplot)
library(phylopomp)
stopifnot(getRversion() >= "4.3")
stopifnot(packageVersion("pomp")>="5.1")
stopifnot(packageVersion("phylopomp")>="0.9.2")
theme_set(theme_bw(base_family="serif"))
set.seed(1159254136)
@

@@

@@

\begin{abstract}

\end{abstract}

@@

\section{Mathematical definitions}

\subsection{Population processes}

@@

\begin{figure}
  \begin{center}
    \resizebox{0.9\linewidth}{!}{\input{figs/model_diagrams}}
  \end{center}
  \caption{
    Examples of compartmental models.
    Demes are shaded.
    \AAK{[Perhaps another one or two examples here?]}
    \AAK{[We could add dots to the deme compartments to signify individuals\dots.]}
    \label{fig:examples}
  }
\end{figure}

@@

\begin{figure}
  \begin{center}
    \resizebox{0.9\linewidth}{!}{
      \begin{tikzpicture}[scale=1]
        \usetikzlibrary{shapes,arrows,positioning}
        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
        \tikzstyle{state}=[shape=ellipse, color=black, draw, font=\large, fill=white, thick, minimum size=5em]
        \tikzstyle{trans}=[color=black, thick, >=stealth]
        \node[state] (s0) at (0,0) {$x=(S,E,I,R)$};
        \node[state] (s1) at (165:7.8) {$x'=(S-1,E+1,I,R)$};
        \node[state] (s2) at (195:7.1) {$x'=(S,E-1,I+1,R+1)$};
        \node[state] (s3) at (295:4) {$x'=(S,E,I-1,R+1)$};
        \node[state] (s4) at (355:9) {$x'=(S+1,E,I,R-1)$};
        \node[state] (s5) at (18:6) {$x'=(S,E,I,R)$};
        \draw[trans,->] (s0) -- (s1) node[midway,above,sloped] {$\mathrm{Trans}$};
        \draw[trans,->] (s0) -- (s2) node[midway,above,sloped] {$\mathrm{Prog}$};
        \draw[trans,->] (s0) -- (s3) node[midway,above,sloped] {$\mathrm{Recov}$};
        \draw[trans,->] (s0) -- (s4) node[midway,above,sloped] {$\mathrm{Wane}$};
        \draw[trans,->] (s0) -- (s5) node[midway,above,sloped] {$\mathrm{Sample}$};
        \node[font=\Large] (U) at ($(s0)+(-6,-5)$) {$\Jumps=\{\mathrm{Trans},\mathrm{Prog},\mathrm{Recov},\mathrm{Wane},\mathrm{Sample}\}$};
      \end{tikzpicture}
    }
  \end{center}
  \caption{
    Markov state transition diagram for an SEIR model.
    The state is characterized by four numbers, $S$, $E$, $I$, and $R$.
    From a given state $x$, there are five possible kinds of events $x\mapsto{x'}$ as shown.
    From the point of view of the induced genealogy process,
    $\mathrm{Trans}$ (transmission) is of birth type,
    $\mathrm{Prog}$ (progression) is of migration type,
    and $\mathrm{Recov}$ (recovery) is of death type,
    while $\mathrm{Wane}$ (loss or waning of immunity) is of neutral type.
    Note that, in this formulation, when a $\mathrm{Sample}$ (sampling) event occurs, the state does not change.
    \label{fig:markov_state}
  }
\end{figure}

\paragraph{Mathematical notation}

Denote the underlying probability space by $(\Omega,\Borel,\mathrm{Prob})$.
Throughout the paper, we will adopt the convention that a bold-face symbol (\eg $\rdm{X}$), denotes a random element while the same symbol in regular face ($X$) denotes an element in its range.
We will assume that our population process is a time-inhomogeneous Markov jump process, $\X_t$, parameterized by time $t\in\Rp\colonequals\{t\in\mathbb{R}\;\vert\;t\ge{0}\}$) and taking values in some space $\Xspace$.
In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
The population process is completely specified by its initial-state distribution, $p_0$, and its transition rates $\alpha$.
In particular, we suppose that
\begin{equation}\label{eq:ic}
  \Prob{\X_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
\end{equation}
for all measurable sets $\mathcal{E}\subseteq\Xspace$.
For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
More precisely, the transition rates have the following properties:
\begin{equation*}
  \begin{gathered}
    \alpha(t,x,x')\ge{0}, \qquad \int_{\Xspace}{\alpha(t,x,x')\,\dd{x'}}<\infty,\\
  \end{gathered}
\end{equation*}
for all $t\in\Rp$ and $x,x'\in\Xspace$.
Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
Let $\N_t$ be the number of jumps that $\X$ has taken by time $t$.
We assume that $\N_t$ is a simple counting process so that
\begin{equation*}
  \begin{gathered}
    \Prob{\N_{t+\Delta}=n+1\;\vert\;\N_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),\\
    \Prob{\N_{t+\Delta}>n+1\;\vert\;\N_{t}=n}=o(\Delta),\\
    \Prob{\X_{t+\Delta}\in\mathcal{E}\;\vert\;\X_{t}=x, \N_{t+\Delta}-\N_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}.
  \end{gathered}
\end{equation*}
We will further assume that $\X_t$ is non-explosive, \ie, that $\Prob{\N_t<\infty}=1$ for all $t$.
\AAK{[Is this equivalent to non-explosivity? Or merely an implication?]}

The above may be compactly summarized by stating that if $w(t,x)$ satisfies the Kolmogorov forward equation (KFE),
\begin{equation}\label{eq:kfe}
  \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},
\end{equation}
and if, moreover, $w(0,x) = p_0(x)$,
then $\int_{\mathcal{E}}{w(t,x)\,\dd{x}}=\Prob{\X_t\in\mathcal{E}}$ for every measurable $\mathcal{E}\subseteq{\Xspace}$.
\Cref{eq:kfe} is sometimes called the \emph{master equation} for $\X_t$.

Without loss of generality, one can assume, as we do here, that the sample paths $t\mapsto\X_t(\omega)$ for $\omega\in\Omega$ are right-continuous with left limits (\ie \cadlag).
In fact, all of the processes we will describe in this paper will be taken to have \cadlag\ sample paths, and we will frequently need to refer to their left-limits.
Accordingly, if $\rdm{Z}_t$ is any \cadlag\ random process, we define
\begin{equation*}
  \leftlim{\rdm{Z}}_t\colonequals
  \begin{cases}
    \lim_{t'\uparrow{t}}\rdm{Z}_t, & t>0,\\
    \rdm{Z}_0, & t=0.\\
  \end{cases}
\end{equation*}
Note that $\leftlim{\rdm{Z}}_t$ is thus left-continuous with right limits.
Denote the set of event times of $\rdm{Z}$ by $\event{\rdm{Z}}{t}=\{e\in[0,t]\;\vert\;\leftlim{\rdm{Z}}_e\ne{\rdm{Z}_e}\}$.

\paragraph{Structured populations, demes}

@@

For any $i\in\Demes$, we let $n_i(\X_t)$ denote the number of lineages present in deme $i$ at time $t$, \ie the \emph{occupancy} of deme $i$.
Thus $n(\X_t)\in\Zp^{\Demes}$ is the vector of deme occupancies.

\paragraph{Jump marks}

In the following, it will be useful to break the jumps into distinct categories.
For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
\begin{equation*}
  \alpha(t,x,x')=\sum_u{\alpha_u(t,x,x')}.
\end{equation*}
In \cref{fig:markov_state}, we use the marks to distinguish biologically distinct events.
Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.

Let us define the \emph{jump mark} process, $\U_t$, to be the mark of the latest jump as of time $t$.
As usual, we take the sample paths, $t\mapsto{\U_t(\omega)}$ for $\omega\in\Omega$, to be \cadlag.
Observe that $(\X_t,\U_t)$ is a Markov process but that $\U_t$ is not.

@@

\paragraph{SEIRS model}

\paragraph{SIIR model}

\paragraph{Linear birth-death model}

\paragraph{Moran model and the Kingman coalescent}

\subsection{The history, inventory, and genealogy processes}

\paragraph{History process}

Consider the Markov process $(\X_t,\U_t)$.
We define its \emph{history process}, $\Hist_t$, to be the restriction of the random function $t\mapsto(\X_t,\U_t)$ to the interval $[0,t]$.
Note that $\Hist_t$ is itself a Markov process.
%% Conditional on $\Hist_t$, both $\X_t$ and $\U_t$ are deterministic, as of course are the Markov chains of event times $\rdm{\hat{T}}_k$ and the embedded chains $\rdm{\hat{X}}_k=\X_{\rdm{\hat{T}}_k}$ and $\rdm{\hat{U}}_k=\U_{\rdm{\hat{T}}_k}$.
Conditional on $\Hist_t$, both $\X_t$ and $\U_t$ are deterministic as are the embedded chain, $(\rdm{\hat{X}}_k,\rdm{\hat{U}}_k)$, and the point process of event times $\rdm{\hat{T}}_k$.
(Recall that $\rdm{\hat{X}}_k=\rdm{X}_{\rdm{\hat{T}}_k}$ and $\rdm{\hat{U}}_k=\rdm{U}_{\rdm{\hat{T}}_k}$.)
The non-explosiveness assumption implies that, for every $t$, $\rdm{K}_t\coloneq\big\vert\event{\Hist}{t}\cap[0,t]\big\vert<\infty$ a.s.
The probability measure, $\pi^H$, for $\Hist_t$ can be expressed in terms of these:
\begin{equation*}%\label{eq:Hdens}
  \begin{aligned}
    \pi^{H}(\dd{H_t})
    =&p_{0}(\hat{X}_{0})\,\dd{\hat{X}_0}\,
    \prod_{k=1}^{K_{t}}{\left(\alpha_{\hat{U}_k}\left(\hat{T}_k,\hat{X}_{k-1},\hat{X}_{k}\right)\,\dd{\hat{X}_k}\,\dd{\hat{T}_k}\right)}\\
    &\qquad\times\exp{\left(-\sum_{k=1}^{K_t}{\int_{\hat{T}_{k-1}}^{\hat{T}_k}{\sum_{u}{\int{\alpha_{u}(t',\hat{X}_{k-1},x')\,\dd{x'}}}}\,\dd{t'}}\right)}.
  \end{aligned}
\end{equation*}

\paragraph{Inventories}

Our goal in this paper is to probabilistically characterize how the genealogical relationships among lineages evolve through time.
Accordingly, we develop some notation for this purpose.
To begin with, we assign to each lineage a unique number $j\in\Zp$.
This can be done in any fashion, so long as no two lineages ever receive the same number.
For example, when a new lineage arises, we can assign it the smallest integer that has not yet been assigned.
We will define the \emph{inventory process}, $\Inv_t$, so that, for every lineage $j\in\Zp$, $\Inv_t(j)$ is the deme in which $j$ is found.
However, when $t$ is before the birth or after the death of lineage $j$, then clearly $\Inv_t(j)\notin\Demes$.
We say in this case that lineage $j$ is in the \emph{underdeme}, which we denote using the symbol $\eth$, so that we can write $\Inv_t(j)=\eth$.
We define $\Demesplus\colonequals\Demes\cup\{\eth\}$ so that $\Inv:\Rp\times\Zp\to\Demesplus$.

The birth and death times of lineage $j$ are therefore
\begin{equation*}
  \begin{gathered}
    t^b_j=\min\{t|\Inv_t(j)\ne{\eth}\}\qquad
    \text{and}\qquad
    t^d_j=\sup\{t|\Inv_t(j)\ne{\eth}\},
  \end{gathered}
\end{equation*}
respectively.
Observe that $n_i(\X_t)=\left\vert\left\{j\;\vert\;\Inv_t(j)=i\right\}\right\vert$ for all $t\in\Rp$ and $i\in\Demes$.
Note also that $n$ does not count the inhabitants of the underdeme.

\paragraph{Jump types}

Different kinds of events that occur for the population process can have different kinds of effects on the inventory process, and indeed not every jump affects $\Inv_t$ at all.
From the point of view of the inventory process, there are five distinct \emph{pure types} of jumps, which we enumerate here.
\begin{enumerate}[(a)]
\item Birth-type events result in the branching of one or more new lineages, each from some existing lineage.
  If $j$ is one of the new lineages, we use the expression $\Anc(j)$ to refer to its ancestor.
  Examples of birth-type events include transmission events, speciations, and actual births.
  It is not assumed that all new lineages arising from a birth event share the same ancestor.
\item Death-type events result in the extinction of one or more lineages.
  Examples include recovery, death of a host, and species extinctions.
\item Migration-type events result in the movement of a lineage from one deme to another.
  Spatial movements, changes in host age or behavior, and progression of an infection can all be represented as migration-type events.
\item Sample-type events result in the collection of a sample from a lineage but do not in themselves affect the inventory process.
\item Neutral-type events result in no change to any of the lineages.
\end{enumerate}
\Cref{fig:markov_state} depicts an example with all five of the pure types.
It is not necessary that a jump fall into just one of these types.
It is allowable, for instance to have compound jumps that fall into more than one category.
For example, sample/death-type events, in which a lineage is simultaneously sampled and removed, have been used, as have birth/death events in which one lineage reproduces at the same moment that another dies.
The theory presented here places no restrictions on the complexity of the events that can occur.

However, we do impose the restriction that the \emph{production}, \ie the deme-specific number of lineages emerging from the event, be constant for all jumps of a given mark.
To be precise, the \emph{production} is defined to be a function $r:\Jumps\times\Demes\to\Zp$, such that $r^u_i$ lineages of deme $i$ emerge from each event of mark $u$.
We write $r^u=(r^u_i)_{i\in\Demes}$.
Note that we lineages that die as a result of an event do not count in the production.
Also, it is important to note that the parent lineage or lineages, if they survive the event, are always counted in the production.

Because different kinds of events may differ not only in the number of offspring they engender, but also in the number of parent lineages, and the distribution of offspring among parents and demes, there is implicitly a deterministic indicator function $Q_u$, for $u\in\Jumps$, (described below) that captures these properties.

\paragraph{Inventory process}

The structure of the state space for the inventory process, $\Inv_t$, has already been described.
It remains to define its stochastic dynamics.
The $\Inv_t$ process is driven by the population process $\X_t$ in that jumps in $\Inv_t$ do not occur except when jumps in $\X_t$ occur:
$\leftlim{\Inv}_t\ne\Inv_t$ implies $\leftlim{\N}_t\ne\N_t$.

At jumps of birth type and mark $u$, the appropriate number of random parents are selected from the appropriate deme(s) and each one sires the appropriate number of offspring in each deme.
At jumps of death type, the appropriate number of lineages are selected from the appropriate demes and moved to the underdeme.
At jumps of migration type, randomly selected lineages are moved between demes as appropriate.
At sample-type and neutral-type jumps, no change occurs.
Jumps of compound type are handled in the obvious way.

Crucially, the assumption that the population process $\X_t$ is Markov empowers us to assume that the individual lineages within each deme are exchangeable with respect to the inventory process.
Since the law governing $\X_t$ is independent of the identities of the individual lineages, and since the individual lineages are exchangeable by assumption, it follows that, conditional on $\Hist_t$, the jumps of the inventory are independent of one another.

\paragraph{Genealogies}

\citet{King2022}, showed how an unstructured population process induces a process on the space of genealogies.
Although we now treat a more general case, the construction is much the same, so we abbreviate the presentation.
Readers wishing for more detail should consult the earlier paper.

>>>>

\documentclass[10pt,reqno,final]{amsart}
\input{header}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,decorations,calc,math}

@@

<<prefix,include=FALSE,cache=FALSE,purl=FALSE>>=
prefix <- "sgp"
source("setup.R")
@
<<packages,include=FALSE,cache=FALSE>>=
library(tidyverse)
library(ggtree)
library(pomp)
library(cowplot)
library(phylopomp)
stopifnot(getRversion() >= "4.3")
stopifnot(packageVersion("pomp")>="5.1")
stopifnot(packageVersion("phylopomp")>="0.9.2")
theme_set(theme_bw(base_family="serif"))
set.seed(1159254136)
@

@@

\begin{abstract}
  We derive an algorithm for the exact computation of the likelihood of an observed transmission tree.
\end{abstract}

@@

@@

\section{Mathematical preliminaries}

\subsection{Notation}

Throughout the paper, we will adopt the convention that a bold-face symbol (\eg $\Xr$), denotes a random element.
We will be concerned with a variety of stochastic processes, in both discrete and continuous time.
In both cases, we will use a subscript to indicate the time parameter: \eg $\Xr_t$ or $Gr_k$, where $t$ takes values in the non-negative reals $\Rp$ and $k$ in the non-negative integers $\Zp$.
In the case of continuous-time processes, we will assume that sample paths are right-continuous with left limits (\ie \cadlag).
We will frequently need to refer to the left-limit of such a process.
Accordingly, if $\Zr_t$ is a \cadlag\ random process, we define
\begin{equation*}
  \Zrt_t\colonequals
  \begin{cases}
    \displaystyle\lim_{t'\,\uparrow\,{t}}\;\Zr_{t'}, & t>0,\\[2ex]
    \Zr_0, & t=0.\\
  \end{cases}
\end{equation*}
Note that $\Zrt_t$ is thus left-continuous with right limits.

If $Zr_t$, $t\in\Rp$ is a pure jump process, knowledge of its sample path is equivalent to knowledge of the number, $\Kr_t$, of jumps it has taken as of time $t$, the jump times $\Trh_k$, and the embedded chain $\Zrh_k\coloneq{\Zr_{\Trh_k}}$, $k=0,\dots,\Kr_t$.
In particular, if we adopt the convention that $\Trh_0=0$ and $\Trh_{\Kr_t+1}=t$, then
$\Zr_t=\Zrh_k$ for $t\in\halfopen{\Trh_k,\Trh_{k+1}}$, $k=0,\dots,\Kr_t$.
If $\Zr$ is such a pure jump process, let $\event{\Zr}{t}=\Set{\Trh_1,\dots,\Trh_{\Kr_t}}$ denote the set of its event times in the interval $[0,t]$.

\subsection{Population process}

@@

\begin{figure}
  \begin{center}
    \resizebox{0.9\linewidth}{!}{
      \begin{tikzpicture}[scale=1]
        \tikzstyle{box}=[color=black, draw, fill=white, very thick, minimum size=3em]
        \tikzstyle{label}=[font=\Large]
        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
        \tikzstyle{flow}=[color=black, very thick, >=stealth]
        \tikzstyle{deme}=[fill=black!20!white]
        \tikzstyle{modulate}=[color=blue, thick, >=latex]
        \coordinate (origin) at (0,0);
        \node[label] (lab) at ($(origin)+(-0.5,2.1)$) {\textbf{A}};
        \input{figs/seir_diagram}
        \coordinate (origin) at (9,0);
        \node[label] (lab) at ($(origin)+(-0.5,2.1)$) {\textbf{B}};
        \input{figs/siir_diagram}
        \coordinate (origin) at (0,-4);
        \node[label] (lab) at ($(origin)+(-0.5,2.1)$) {\textbf{C}};
        \input{figs/covid_diagram}
      \end{tikzpicture}
    }
  \end{center}
  \caption{
    Examples of discretely-structured population models.
    Demes are shaded.
    (A) An SEIRS model.
    Susceptible individuals ($\lab{S}$), once infected, enter a transient incubation phase ($\lab{E}$) before they become infectious ($\lab{I}$).
    Upon recovery ($\lab{R}$), individuals experience immunity from reinfection.
    If this immunity wanes, they re-enter the susceptible compartment.
    Pathogen lineages are to be found in hosts within the $\lab{E}$ and $\lab{I}$ compartments only.
    Accordingly, there are two demes ($\Demes=\Set{\lab{E},\lab{I}}$).
    If there is exactly one lineage per host, then the occupancy, $n(\Xr_t)=(n_{\lab{E}}(\Xr_t),n_{\lab{I}}(\Xr_t))$, is the integer 2-vector giving the numbers of hosts in the respective compartments.
    \AAK{[Perhaps another one or two examples here?]}
    \AAK{[We could add dots to the deme compartments to signify individuals\dots.]}
    \label{fig:examples}
  }
\end{figure}

We will assume that our population process is a time-inhomogeneous Markov jump process, $\Xr_t$, $t\in\Rp$, taking values in some space $\Xspace$.
In earlier work \citep{King2022}, we limited ourselves to the case $\Xspace=\Z^d$, but here we assume only that $\Xspace$ is a complete metric space with a countable dense subset, \ie a Polish space.
The population process is completely specified by its initial-state distribution, $p_0$, and its transition rates $\alpha$.
In particular, we suppose that
\begin{equation}\label{eq:ic}
  \Prob{\Xr_0\in\mathcal{E}}=\int_{\mathcal{E}}{p_0(x)\,\dd{x}}
\end{equation}
for all measurable sets $\mathcal{E}\subseteq\Xspace$.
For any $t\in\Rp$, $x,x'\in\Xspace$, we think of the quantity $\alpha(t,x,x')$ as the instantaneous hazard of a jump from $x$ to $x'$.
More precisely, the transition rates have the following properties:
\begin{equation*}
  \begin{gathered}
    \alpha(t,x,x')\ge{0}, \qquad \int_{\Xspace}{\alpha(t,x,x')\,\dd{x'}}<\infty,\\
  \end{gathered}
\end{equation*}
for all $t\in\Rp$ and $x,x'\in\Xspace$.
Henceforth, we understand that integrals are taken over all of $\Xspace$ unless otherwise specified.
Let $\Kr_t$ be the number of jumps that $\Xr$ has taken by time $t$.
We assume that $\Kr_t$ is a simple counting process so that
\begin{equation*}
  \begin{gathered}
    \CondProb{\Kr_{t+\Delta}=n+1}{\Kr_{t}=n}=\Delta\,\int{\alpha(t,x,x')\,\dd{x'}}+o(\Delta),\\
    \CondProb{\Kr_{t+\Delta}>n+1}{\Kr_{t}=n}=o(\Delta),\\
    \CondProb{\Xr_{t+\Delta}\in\mathcal{E}}{\Xr_{t}=x, \Kr_{t+\Delta}-\Kr_{t}=1}=\frac{\int_{\mathcal{E}}{\alpha(t,x,x')\,\dd{x'}}}{\int{\alpha(t,x,x')\,\dd{x'}}}.
  \end{gathered}
\end{equation*}
We will further assume that $\Xr_t$ is non-explosive, \ie that $\Prob{\Kr_t<\infty}=1$ for all $t$.

\subsubsection{Kolmogorov forward equation}

The above may be compactly summarized by stating that if $w(t,x)$ satisfies the Kolmogorov forward equation (KFE),
\begin{equation}\label{eq:kfe}
  \frac{\partial{w}}{\partial{t}}(t,x)=\int\!w(t,x')\,\alpha(t,x',x)\,\dd{x'}-\int\!w(t,x)\,\alpha(t,x,x')\,\dd{x'},
\end{equation}
and if, moreover, $w(0,x) = p_0(x)$,
then $\int_{\mathcal{E}}\!{w(t,x)\,\dd{x}}=\Prob{\Xr_t\in\mathcal{E}}$ for every measurable $\mathcal{E}\subseteq{\Xspace}$.
\Cref{eq:kfe} is sometimes called the \emph{master equation} for $\Xr_t$.

@@

\begin{figure}
  \begin{center}
    \resizebox{0.9\linewidth}{!}{
      \begin{tikzpicture}[scale=1]
        \usetikzlibrary{shapes,arrows,positioning}
        \tikzstyle{coordinate}=[inner sep=0pt,outer sep=0pt]
        \tikzstyle{state}=[shape=ellipse, color=black, draw, font=\LARGE, fill=white, thick, minimum size=5em]
        \tikzstyle{trans}=[color=black, thick, >=stealth]
        \node[state] (base) at (0,0) {$x=(S,E,I,R)$};
        \node[state] (trans) at (165:8.4) {$x'=(S-1,E+1,I,R)$};
        \node[state] (prog) at (195:8) {$x'=(S,E-1,I+1,R+1)$};
        \node[state] (recov) at (295:4) {$x'=(S,E,I-1,R+1)$};
        \node[state] (wane) at (355:9) {$x'=(S+1,E,I,R-1)$};
        \node[state] (sample) at (18:7.5) {$x'=(S,E,I,R)$};
        \draw[trans,->] (base) -- (trans) node[midway,above,sloped] {$\lab{Trans}$};
        \draw[trans,->] (base) -- (prog) node[midway,above,sloped] {$\lab{Prog}$};
        \draw[trans,->] (base) -- (recov) node[midway,above,sloped] {$\lab{Recov}$};
        \draw[trans,->] (base) -- (wane) node[midway,above,sloped] {$\lab{Wane}$};
        \draw[trans,->] (base) -- (sample) node[midway,above,sloped] {$\lab{Sample}$};
        \node[font=\Large] (U) at (95:3.2) {$\Jumps=\Set{\lab{Trans},\lab{Prog},\lab{Recov},\lab{Wane},\lab{Sample}}$};
      \end{tikzpicture}
    }
  \end{center}
  \caption{
    Markov state transition diagram for the SEIRS model depicted in \cref{fig:examples}A.
    The state, $x$, is characterized by four numbers, $S$, $E$, $I$, and $R$.
    From a given state $x$, there are five possible kinds of jumps $x\mapsto{x'}$.
    Accordingly, the set, $\Jumps$, of jump marks has five elements.
    Each of these is of a different type:
    $\lab{Trans}$ (transmission) is of birth type,
    $\lab{Prog}$ (progression) is of migration type,
    $\lab{Recov}$ (recovery) is of death type,
    $\lab{Sample}$ (sampling) is of sample type,
    and $\lab{Wane}$ (loss or waning of immunity) is of neutral type.
    See \cref{sec:jump_types} for a description of these jump types.
    Note that, in this formulation, when a sampling event occurs, the state does not change.
    \label{fig:markov_state}
  }
\end{figure}

\subsubsection{Structured populations, demes}

@@

For any $i\in\Demes$, we let $n_i(\Xr_t)$ denote the number of lineages present in deme $i$ at time $t$, \ie the \emph{occupancy} of deme $i$.
Thus $n(\Xr_t)\in\Zp^{\Demes}$ is the vector of deme occupancies.

\subsubsection{Jump marks}

In the following, it will be useful to break the jumps into distinct categories.
For this purpose, we let $\Jumps$ be a countable set of jump \emph{marks} such that
\begin{equation*}
  \alpha(t,x,x')=\sum_u{\alpha_u(t,x,x')}.
\end{equation*}
In \cref{fig:markov_state}, we use the marks to distinguish events of different types.
Here and in the following, sums over $u$ are taken over the whole of $\Jumps$ unless otherwise indicated.

Let us define the \emph{jump mark} process, $\Ur_t$, to be the mark of the latest jump as of time $t$.
As usual, we take the sample paths of $\Ur_t$ to be \cadlag.
Observe that $(\Xr_t,\Ur_t)$ is a Markov process, though $\Ur_t$ is not.

@@

\subsubsection{SEIRS model}

\subsubsection{SIIR model}

\subsubsection{Linear birth-death model}

\subsubsection{Moran model and the Kingman coalescent}

\subsection{History process}

Consider the Markov process $(\Xr_t,\Ur_t)$.
We define its \emph{history process}, $\Hr_t$, to be the restriction of the random function $s\mapsto(\Xr_s,\Ur_s)$ to the interval $[0,t]$.
Note that $\Hr_t$ is itself trivially a Markov process, since it contains its own history.
Alternatively, one can think of $\Hr_t$ as consisting of the sequence
$\left(\left(\Trh_k,\Xrh_k,\Urh_k\right)\right)_{k=0}^{\Kr_t}$.
In particular, conditional on $\Hr_t$, both $\Xr_t$ and $\Ur_t$ are deterministic as are $\Kr_t$ and the embedded chains, $\Xrh_k$, $\Urh_k$, and the point process of event times $\Trh_k$.
The probability measure, $\pi^{\mathrm{H}}$, for $\Hr_t$ can be expressed in terms of these:
\begin{equation*}\label{eq:Hdens}
  \begin{aligned}
    \pi^{\mathrm{H}}(\dd{\H_t})
    =&p_{0}(\Xh_{0})\,\dd{\Xh_0}\,
    \prod_{k=1}^{K_{t}}{\alpha_{\Uh_k}\!\!\left(\Th_k,\Xh_{k-1},\Xh_{k}\right)\,\dd{\Xh_k}\,\dd{\Th_k}}\\
    &\qquad\times\exp{\left(-\sum_{k=0}^{K_t}{\int_{\Th_{k}}^{\Th_{k+1}}{\sum_{u}{\int{\alpha_{u}(t',\Xh_{k},x')\,\dd{x'}}}}\,\dd{t'}}\right)},
  \end{aligned}
\end{equation*}
where again, by convention, $\Th_0=0$ and $\Th_{K_t+1}=t$.

\subsection{Binomial ratio}\label{sec:binomial_ratio}

For $n,r,\ell,s\in{\Zp^\Demes}$, define the \emph{binomial ratio}
\begin{equation*}
  \BinRatio{n}{\ell}{r}{s}\colonequals
  \begin{cases}
    \frac{\displaystyle\prod_{i\in\Demes}{\binom{n_i-\ell_i}{r_i-s_i}}}%
         {\displaystyle\prod_{i\in\Demes}{\binom{n_i}{r_i}}},
         & \text{if}\ \forall i\ n_i\ge{\Set{\ell_i,r_i}}\ge{s_i}\ge{0},\\
         0, & \text{otherwise}.
  \end{cases}
\end{equation*}
Observe that $\BinRatio{n}{\ell}{r}{s}\in{[0,1]}$.
Moreover, in consequence of the Chu-Vandermonde identity, we have
\begin{equation*}
  \sum_{s\in\Zp^{\Demes}}\BinRatio{n}{\ell}{r}{s}\binom{\ell}{s}=1,
\end{equation*}
whenever $n_i\ge{\Set{\ell_i,r_i}}\ge{0}$ for all $i$.

\section{The induced genealogy process}

\subsection{Genealogies}

For our purposes, a \emph{genealogy}, $G_{T}$, is a labeled, time-calibrated tree.
It encapsulates the relationships of shared ancestry among a set of lineages that are alive at time $T\in\Rp$, and perhaps a set of samples collected at earlier times.
There are three distinct kinds of nodes:
\begin{inparaenum}[(i)]
\item \emph{tip nodes}, which represent labeled living lineages;
\item \emph{internal nodes}, which represent events at which lineages diverged and/or moved from one deme to another; and
\item \emph{sample nodes}, which represent labeled samples.
\end{inparaenum}
Each node $a$ is associated with a specific time, $t(a)$.
In particular, if $a$ is a tip node in $G_T$, then $t(a)=T$;
if $a$ is a sample node, then $t(a)$ is the time at which the sample was taken.
Moreover, if node $a$ is ancestral to node $a'$, then $t(a)\le{t(a')}$ and $t(a')-t(a)$ is the distance between $a$ and $a'$ along the genealogy.
We let $\event{G}{T}$ denote the set of all node-times of the genealogy $G_T$.
Importantly, a genealogy informs us not only about the shared ancestry of any pair of lineages, but also about where in the set of demes any given lineage was at all times.
Accordingly, we can visualize a genealogy as a tree, the nodes and edges of which are painted, with a distinct color for each deme.

Formally, we define a genealogy, $G_T$, to be an ordered pair $(y,z)$.
The $z$ element specifies the topological structure, as follows.
Let $\leaves{G_T}$ be the set of all tip and sample nodes in $G_T$ and define $\func{P}$ to be the set of all partitions of $\leaves{G_T}$.
The partition \emph{fineness} defines a partial order on $\func{P}$.
Specifically, for $p,p'\in{\func{P}}$, we say $p\preceq{p'}$ if and only if for every $c'\in{p'}$ there is a unique $c\in{p}$ such that $c\supseteq{c'}$.
The topological structure of $G_T$ is defined by a map $z:[0,T]\to\func{P}$ that is \cadlag\ and monotone in the sense that $t_1<t_2$ implies $\z_{t_1}\preceq{\z_{t_2}}$.

The $y$ element of $G_T$ specifies the coloring of branches and location of nodes.
In particular, for $a\in\leaves{G_T}$ and $t\in[0,T]$, define $d_{t}(a)$ to be the deme in which the lineage of $a$ was found at time $t$.
To make this well-defined, let $d_{t}(a)=\eth$ for $t\in{[t(a),T]}$;
we call $\eth$ the \emph{underdeme} and define $\Demesplus\coloneq{\Demes\cup\{\eth\}}$.
Thus $d:[0,T]\times\leaves{G_T}\to\Demesplus$ is a well-defined \cadlag\ function.

Since not every node along a lineage corresponds to a change of deme, we also count the number of nodes along each lineage.
Let $m:[0,T]\times\leaves{G_T}\to\Zp$ be such that $m_{t}(a)$ is the number of internal or sample nodes encountered along lineage $a$ as one goes from time $0$ to time $t\le{T}$ along the genealogy.
For $t\in{[t(a),T]}$, let $m_{t}(a)=m_{t(a)}(a)$.
Clearly, $m_t(a)$ is a simple counting process, with $m_0(a)=0$ for all $a$.

Finally, we define the map $y:[0,T]\times\leaves{G_T}\to\Demesplus\times\Zp$ by $y_{t}(a)=\left(d_{t}(a),m_{t}(a)\right)$.
The two \cadlag\ functions $y$ and $z$ specify the genealogy $G_T$ completely and uniquely.
When the genealogy is not clear from context, we write $y^{G}$ and $z^{G}$ to refer to the respective components of genealogy $G$.

Clearly, the topological structure of a tree imposes constraints on the permissible functions $y$ that describe the painting and node placement along it.
In particular, $y$ must respect the topological structure of $G_T$, such that if two lineages share a common ancestor at time $t$, both lineages must have been in the same deme at that time and the number of previous events along the lineage must agree.
In other words, if $a,b\in{c}\in{p}\in{\z_{t}}$, then $y_{t}(a)=y_{t}(b)$.
Let $\Yspace^z$ denote the space of all functions that respect the topological structure described by $z$.
Let $\Yspace^\z_t$ be the time-$t$ \emph{section} of $\Yspace^z$, \ie $\Yspace^\z_t=\displaystyle\bigcup_{y\in\Yspace^z}{\Set{y_t}}$.

\subsection{Jump types}\label{sec:jump_types}
